\hspace{1cm} Do syntactic and semantic similarity lead to interference effects?\newline Evidence from self-paced reading and event-related potentials using German

### 2023-10-15 paper/main.tex
minor edits

Considered together, the present SPR and ERP experiments consistently showed effects of semantic interference and no effects of syntactic interference. This pattern was not only found in the primary dependent measures of the experiments, respectively, but also in the comprehension accuracy. 
The modeling reported here suggests that one way to explain the observed pattern in the context of cue-based retrieval accounts is by assuming that the parser intelligently searches for a subject that is within the same clause. A broader implication is that the human sentence comprehension system may be using the hierarchical syntactic structure to selectively target nouns for retrieval; this is an idea that has independent support in the literature \cite[e.g.,][]{Sturt2003,dillon2013}.

@@

In the remainder of this section, we discuss several important issues that relate to previous experiments that investigated syntactic and semantic interference.

@@

Thus, the words directly preceding the critical verb did not show the semantic interference effect. A further question that arises is: which brain response was elicited by the distractor itself (see Figure \ref{fig:erp_distractor})? An exploratory visual inspection of the ERPs elicited by the distractor shows several differences between conditions, but no semantic interference effect in the N400 spatio-temporal window. We believe that the differences in the ERPs elicited by the distractor in the different conditions are caused by two factors. First, the high and low syntactic interference conditions have different word orders, i.e., the distractors are embedded in different phrase structures (simple noun phrase vs.\ noun phrase including an adjective within a prepositional phrase; see Example \ref{ex:materials}). This is also apparent from the large difference between high and low syntactic interference conditions before distractor onset. Second, the distractors were not identical across conditions, therefore the ERPs in Figure \ref{fig:erp_distractor} are elicited by different words in the high vs.\ low semantic interference conditions. In summary, this exploratory analysis suggests that the distractors did induce different brain responses across conditions, but these could be due to factors other than encoding intererence. More importantly, these effects did not linger throughout the rest of the sentence (see Figure \ref{fig:erp_precrit}).

@@

The comparison of the results of our SPR and ERP experiments suggests that the SPR results were strongly affected by encoding interference, while the ERPs elicited by the critical verb were not. This difference between methods might be caused by the difference in presentation rate. In the self-paced reading experiment, as implied by the name, the participants had control over the presentation rate of the stimuli. The observed slowdown starting at the distractor  might have resulted from the probably subconscious intent to slow down the sampling rate of the incoming stimuli to gain more time to process them. In contrast, in the ERP experiment, we used rapid serial visual presentation (RSVP), which presents stimuli at a fixed rate that is beyond the control of the participant. Therefore, they must adapt to the set rate and process the stimuli as fast as the rate requires. Here, it is noteworthy that the presentation duration of 500\,ms per word and 400\,ms between words which was used in the present ERP experiment is slower than usual natural reading speed. However, while in SPR the slowdown due to encoding and maintaining the distractor in memory overshadowed potential retrieval effects, the procedure of the ERP experiment prevented that. In the ERP results, we observed an effect at the critical verb, which can be attributed to the retrieval process hypothesized by cue-based retrieval accounts.

@@

The evidence against a main effect of syntactic interference for subject-verb dependency resolution in our study is not consistent with the assumption in previous work, e.g., \cite{vandyke07,mertzen}, that the parser searches for a subject simply by setting the retrieval cue \{± grammatical subject\}.
  However, the semantic interference effect we found can be easily reconciled with previous work in  sentence processing. In the modeling section above, we have already discussed how our results relate to predictions from the \citet{Lewis2005} cue-based retrieval model.  
  %Our results may also be consisted with constraint-based models \citep[e.g.,][]{macdonald1994, mcrae1998modeling, trueswell1993}. Under constraint-based accounts, it is assumed that multiple constraints simultaneously and immediately influence processing. These accounts predict our ERP results: Compared to an inanimate distractor, an animate distractor -- regardless of its syntactic status -- would impede processing at a verb which requires an animate subject. The lack of a syntactic interference effect could indicate that the status `grammatical subject' was not used as a syntactic constraint. As discussed above, a plausible syntactic constraint is that a verb and its arguments need to be within the same clause. 
 Our results could also be seen as consistent with the good-enough processing account \citep{ferreira2007goodenough}, which assumes that comprehenders do not always aim to build a fully fleshed-out analysis of a sentence, but instead might accept incomplete, underspecified, or even incorrect analyses. The high number of participants that needed to be excluded in our experiments due to accuracy below 70\,\% (117 out of 908 participants in the SPR experiment, 29 out of 146 participants in the EEG experiment), suggests that our materials led to high processing demands. It is reasonable to assume that the participants might have--at least occasionally--adopted a good-enough processing mode when faced with high task demands \parencite{swets2008underspecification,LogacevMultiple} or due to working memory capacity limitations \parencite{MalsburgVasishth2013}, or both. So, given a good-enough processing mode which leaves some syntactic relations within the sentence underspecified, it makes sense for the comprehender to use a simple heuristic for subject-verb dependency resolution. Animacy is an obvious choice for such a heuristic. Animate entities are proto-typical agents \citep{dowty1991thematic} and therefore, the primary use of the \{+ animate\} cue to retrieve a subject leads to the correct analysis with high probability (not just within the experiment context, but in everyday language use). Finally, our findings are also consistent with the possibility that language processing relies predominantly on semantic associations to form (probabilistic)
representation of event structures. This view has been put forward by \cite{rabovsky_etal_2018}, when they used the neural-network sentence gestalt model \parencite{mcclelland1989_sentence_gestalt}, to model the N400 amplitude. This model assumes that language comprehension relies on associative form-to-meaning mapping instead of syntactic rules. This assumption is consistent with our results.

Without the modeling results reported above, it would be easy to conclude that our data show that  syntactic cues play no role in sentence processing. However, this conclusion would be obviously wrong. First, as we showed above, the retrieval model suggests that--at least in our data--the subject cue is used, but differently than previously assumed: the relevant cue is subject-in-same-clause. Second, there is plenty of independent evidence for the central role of syntactic information in comprehension; some examples are the role of case marking \parencite{ALV2019,HusainEtAl2014}, syntactic constraints like Principle A of the binding theory \parencite{Sturt2003,dillon2013,yadav2021individual},  and 
the importance of word order \parencite{meng2000mode}. There are of course important examples of the parser ignoring syntactic information. Three dramatic examples are the grammaticality illusion in multiple embeddings \parencite{gibsonthomas99}, the agreement attraction phenomenon \parencite{wagersetal,lago_etal_2021,jaeger_etal_2017}, and local coherence effects \parencite{taboretal04}.  However, there are languages (such as German and Dutch) in which the grammaticality illusion does not occur \parencite{VSLK11,FrankEtAl2015}, and a large-sample study \parencite{lcpaape2023} presents evidence against the syntactic local coherence effect that was reported by \cite{taboretal04}. Thus, it seems that the role of syntax in building incremental structure remains as important as ever.

@@

This project investigated the use of  syntactic and semantic cues during subject-verb dependency resolution with self-paced reading and event-related potentials.  Both our experiments consistently showed semantic interference, i.e., more processing difficulty during long-distance subject-verb dependency resolution in the presence of a semantically matching distractor. This manifested in longer self-paced reading times starting at the distractor and continuing into the post-critical region, and in a more negative N400 amplitude for the high vs.\ low semantic interference conditions elicited by the critical verb. These different time courses in the two experiments are likely to be due to different types of interference: Encoding interference in the SPR experiment and retrieval interference in the ERP experiment. These differences are likely to be due to the methods used (self-paced vs.\ auto-paced reading).Overall, we found evidence against syntactic interference; computational modeling shows that the most likely explanation for this is that the parser uses the syntactic cue  \{± subject-in-same-clause\}  to identify the correct target for retrieval. A broader implication is that the parser may be able to use hierarchical structure in the input to intelligently target the correct syntactic dependent from memory.

>>>>

Taken together, the SPR and ERP experiments consistently showed effects of semantic interference and no effects of syntactic interference. This pattern was not only found in the dependent measures of the experiments that were of primary interest, but also in the comprehension accuracy. 
The modeling reported here suggests that one way to explain the observed pattern in the context of cue-based retrieval is by assuming that the parser intelligently searches for a subject that is within the same clause, and uses the animacy cue without reference to the clause that a noun appears in. A broader implication is that the human sentence comprehension system may be using the hierarchical syntactic structure to selectively target nouns for retrieval; this is an idea that has independent support in the literature \cite[e.g.,][]{Sturt2003,dillon2013}.

@@

In the remainder of this section, we discuss two important issues that relate to previous experiments that investigated syntactic and semantic interference. The first is the role of encoding vs.\ retrieval interference, and the second is broader implications of our findings in the context of previous related work.

@@

Thus, the words directly preceding the critical verb do not show a pattern consistent with semantic interference. A further question that arises is: which brain response was elicited by the distractor itself (see Figure \ref{fig:erp_distractor})? An exploratory visual inspection of the ERPs elicited by the distractor shows several differences between conditions, but no semantic interference effect in the N400 spatio-temporal window. We believe that the differences in the ERPs elicited by the distractor in the different conditions are caused by two factors. First, the high and low syntactic interference conditions have different word orders, i.e., the distractors are embedded in different phrase structures (simple noun phrase vs.\ noun phrase including an adjective within a prepositional phrase; see Example \ref{ex:materials}). This is also apparent from the large difference between high and low syntactic interference conditions before distractor onset. Second, the distractors were not identical across conditions, therefore the ERPs in Figure \ref{fig:erp_distractor} are elicited by different words in the high vs.\ low semantic interference conditions. In summary, this exploratory analysis suggests that the distractors did induce different brain responses across conditions, but these could be due to factors other than encoding intererence. More importantly, these effects did not linger throughout the rest of the sentence (see Figure \ref{fig:erp_precrit}).

@@

The comparison of the results of our SPR and ERP experiments suggests that the SPR results were strongly affected by encoding interference, while the ERPs elicited by the critical verb were not. This difference between methods might be caused by the difference in presentation rate. In the self-paced reading experiment, as implied by the name, the participants had control over the presentation rate of the stimuli. The observed slowdown starting at the distractor  might have resulted from the probably subconscious intent to slow down the sampling rate of the incoming stimuli to gain more time to process them. In contrast, in the ERP experiment, we used rapid serial visual presentation (RSVP), which presents stimuli at a fixed rate that is beyond the control of the participant. Therefore, they must adapt to the set rate and process the stimuli as fast as the rate requires. Here, it is noteworthy that the presentation duration of 500\,ms per word and 400\,ms between words which was used in the present ERP experiment is slower than usual natural reading speed. However, while in SPR the slowdown due to encoding and maintaining the distractor in memory overshadowed potential retrieval effects, the procedure of the ERP experiment might have prevented that. In the ERP results, we observed an effect at the critical verb, which could be attributed exclusively to the retrieval process hypothesized by cue-based retrieval accounts. Having said that, we cannot rule out the possibility that--due to the auto-paced nature of the ERP experiment--any encoding interference that started at or after the distractor is mixed in with the cost of retrieval interference. Indeed, recent modeling work on agreement attraction in reading \parencite{Yadavetal2022} has shown conclusively that some kind of feature distortion as well as cue-based retrieval are needed to explain the benchmark data-sets that exist so far on agreement attraction in reading. If the  \cite{Yadavetal2022} extends to interference effects in general, 
it is reasonable to assume that the effect observed at the verb in the ERP study is driven by both encoding and retrieval interference. However, we cannot resolve the relative roles of these latent processes in the present study.

@@

The evidence against a main effect of syntactic interference for subject-verb dependency resolution in our study is not consistent with the assumption in previous work, e.g., \cite{vandyke07,mertzen}, that the parser searches for a subject simply by setting the retrieval cue \{± grammatical subject\}.
  However, the semantic interference effect we found can be easily reconciled with previous work in  sentence processing. In the modeling section above, we have already discussed how our results relate to predictions from the \citet{Lewis2005} cue-based retrieval model.  
  %Our results may also be consisted with constraint-based models \citep[e.g.,][]{macdonald1994, mcrae1998modeling, trueswell1993}. Under constraint-based accounts, it is assumed that multiple constraints simultaneously and immediately influence processing. These accounts predict our ERP results: Compared to an inanimate distractor, an animate distractor -- regardless of its syntactic status -- would impede processing at a verb which requires an animate subject. The lack of a syntactic interference effect could indicate that the status `grammatical subject' was not used as a syntactic constraint. As discussed above, a plausible syntactic constraint is that a verb and its arguments need to be within the same clause. 
 Our results could also be seen as consistent with the good-enough processing account \citep{ferreira2007goodenough}, which assumes that comprehenders do not always aim to build a fully fleshed-out analysis of a sentence, but instead might accept incomplete, underspecified, or even incorrect analyses. The high number of participants that needed to be excluded in our experiments due to accuracy below 70\,\% (117 out of 908 participants in the SPR experiment, 29 out of 146 participants in the EEG experiment), suggests that our materials led to high processing demands. It is reasonable to assume that the participants might have--at least occasionally--adopted a good-enough processing mode when faced with high task demands \parencite{swets2008underspecification,LogacevMultiple,LogacevVasishthQJEP2016} or due to working memory capacity limitations \parencite{MalsburgVasishth2013}, or both. So, given a good-enough processing mode which leaves some syntactic relations within the sentence underspecified, it makes sense for the comprehender to use a simple heuristic for subject-verb dependency resolution. Animacy is an obvious choice for such a heuristic. Animate entities are proto-typical agents \citep{dowty1991thematic} and therefore, the primary use of the \{+ animate\} cue to retrieve a subject leads to the correct analysis with high probability (not just within the experiment context, but in everyday language use). Finally, our findings are also consistent with the possibility that language processing relies predominantly on semantic associations to form (probabilistic)
representation of event structures. This view has been put forward by \cite{rabovsky_etal_2018}, when they used the neural-network sentence gestalt model \parencite{mcclelland1989_sentence_gestalt}, to model the N400 amplitude. This model assumes that language comprehension relies on associative form-to-meaning mapping instead of syntactic rules. This assumption is consistent with our results.

Without the modeling results reported above, it would be easy to conclude that our data show that  syntactic cues play no role in sentence processing. However, this conclusion would be obviously wrong. First, as we showed above, the retrieval model suggests that--at least in our data--the subject cue is used, but differently than previously assumed: the relevant cue may be subject-in-same-clause. Second, there is plenty of independent evidence for the central role of syntactic information in comprehension; some examples are the role of case marking \parencite[e.g.,][]{ALV2019,HusainEtAl2014,bhatia2022processing,bader2000case,bader2006case,miyamoto2002case}, syntactic constraints like Principle A of the binding theory \parencite[e.g.,][]{Sturt2003,dillon2013,yadav2021individual},  and 
the importance of word order \parencite[e.g.,][]{meng2000mode}. There are of course important examples of the parser ignoring syntactic information. Three dramatic examples are the grammaticality illusion in multiple embeddings in English \parencite{gibsonthomas99}, the agreement attraction phenomenon across different languages \parencite{wagersetal,lago_etal_2021,jaeger_etal_2017}, and local coherence effects \parencite{taboretal04}.  However, there are languages (such as German and Dutch) in which the grammaticality illusion does not occur \parencite{VSLK11,FrankEtAl2015}, and a large-sample study \parencite{lcpaape2023} presents evidence against the syntactic local coherence effect that was reported by \cite{taboretal04}. Also, there seems to be at least one language that has rich case marking (Czech) that does not show any evidence of agreement attraction \parencite{chromy2023number}. Thus, it seems that, cross-linguistically, the role of syntax in building incremental structure and in completing dependencies remains as important as ever.

@@

This project investigated the use of  syntactic and semantic cues during subject-verb dependency resolution with self-paced reading and event-related potentials.  Both our experiments consistently showed semantic interference, i.e., more processing difficulty during long-distance subject-verb dependency resolution in the presence of a semantically matching distractor. This manifested in longer self-paced reading times starting at the distractor and continuing into the post-critical region, and in a more negative N400 amplitude for the high vs.\ low semantic interference conditions elicited by the critical verb. These different time courses in the two experiments are likely to be due to different types of interference: Encoding interference in the SPR experiment and retrieval (and possibly also encoding) interference in the ERP experiment. These differences in the time course are likely to be due to the methods used in the two sets of experiments (self-paced vs.\ auto-paced reading). Overall, we found evidence against the use of a syntactic cue such as \{$\pm$ grammatical subject\}, as previously assumed in the literature; computational modeling shows that the most likely explanation for the observed data is that the parser uses the syntactic cue  \{± subject-in-same-clause\}  to identify the correct target for retrieval.  A broader implication is that the parser may be able to use hierarchical structure in the input to intelligently target the correct syntactic dependent from memory, leading to no interference from a distractor that is in subject position within an embedded clause.
